Firstly, I would would like to express my gratitude to my supervisors,
Prof.~Pat Morin, Prof.~Vida Dujmovi\'{c}, and Prof.~``Jit'' Bose, for
welcoming me to Ottawa and Carleton University, and helping me work on
and study other topics before settling on the study of encoding
arguments as the subject of my thesis. I am especially grateful to
Prof.~Morin, whose insight, knowledge, and personal supervision have
been absolutely invaluable.

I am also indebted to the examiners of my thesis, Prof.~Jean-Lou De
Carufel and Prof.~Daniel Panario, for their meticulous and careful
reading of my work, and for their helpful comments.

Thanks to Prof.~Luc Devroye at McGill University for teaching a
wonderful class in the study of algorithms and data structures, which
sparked my interest in information theory.

Of course, I would like to thank my fellow labmates and researchers
working in the Computational Geometry Lab at Carleton University,
whose company I found to be appropriately motivating whenever I felt
inspired, and appropriately distracting whenever I felt overwhelmed. I
am convinced that the rare atmosphere sustained in our lab is
precious.

Finally, I thank my family and friends, since without them I would be
nowhere. I am most grateful to my parents who have supported me
throughout all of my studies and all of my life. A thousand thanks go
out to my girlfriend Emily, for her support and encouragement. Special
thanks go out to my friend Costin B\u{a}descu, who drew me into
research and the study of mathematics and computer science. I would
also like to thank my friends Ash Chapman, Troy Hildebrandt, and
Darryl Hill, whose solidarity as fellow students has been a great
comfort. I am glad that we could confront the looming presence of
graduate research together. And lastly, thanks to Bavaria Brewery, for
offering the most drinkable cheap beer in Ottawa.

%I recognize that this research would not have been possible without
%financial assistance from NSERC and Carleton University.